Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-09-09 bmc_template/manuscript.tex
almost-last pass on MS?

@@

We used the {\tt depmixS4} package
to fit covariate-dependent transition HMMs, 
simulate states and step lengths using the
estimated parameters, and estimate the most likely
states with the Viterbi
algorithm.

@@

Model complexity and the number of parameters increase as the number of latent states increase.
For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and finally by FMMs and HMMs incorporating temporal heterogeneity. In general, the number of free parameters in an HMM is the sum of the number of free parameters for each of the three model components. Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$. Then the number of free parameters of an HMM is:
$k_{i}\cdot (n-1)$ (\emph{Initial states})
$+ k_{t}\cdot n\cdot (n-1)$ (\emph{Transition probabilities})
$+ k_{e}\cdot n$ (\emph{Emission parameters}).
As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs increases linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically. When comparing BICs, it is important to account for the tradeoff between log-likelihood and number of states, but also log-likelihood and number of free parameters.

We used simulations to predict expected hourly step lengths and
autocorrelation functions (ACF).  While the computation of expected
step length and ACF is straightforward for FMMs, and feasible for
homogeneous HMMs, the interaction between the geometric dwell time
within each state and the temporally varying interaction probabilities
makes it infeasible for more complex models.  We used this approach to
validate our models, comparing our simulated predictions with the
observed movements.  The more usual approach, generating predictions
from the expected step lengths conditional on the most likely state
sequence predicted by the Viterbi algorithm or pseudo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is somewhat problematic
because the predictions by these methods rely on the observed data. 
This approach is useful to predict
missing data in the observation sequence, but because it is
conditional on the observed values, it can not reliably evaluate
goodness of fit for HMM models with
different degrees of structural complexity.

@@

%% fig 1 goes somewhere here
The simulation supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states. Heterogeneous transition models can always predict the correct number of states, whereas the temporally homogeneous models overestimate the number of states (based on BIC-optimality: Figure 1).

\bmb{need to describe in words what the figure says}

@@

The BIC-optimal number of states for time homogeneous models is consistent with van de Kerk et al.'s \cite{kerk2015hidden} results. For time homogeneous models, the Weibull-wrapped-Cauchy \cite{kerk2015hidden}, Weibull von Mises, and log Normal without turning angles all identify the same BIC-optimal number of states. Furthermore, the homogeneous-HMM models do vary among models (different step-length distributions, with/without turning angles) in how many states they identify, whereas, the heterogeneous-HMM models are consistent among models (Figure 2).

@@

Models with temporal heterogeneity are better (lower BIC) than homogeneous models in both FMM and HMM frameworks, but time-homogeneous HMMs are better than FMMs with sinusoidal temporal heterogeneity (Figure 3). Turning to the temporally heterogeneous HMMs (Figure 3, right panel), we see that the model with different transition probabilities for each hour of the day (HMM + THhourly) is overparameterized; it underperforms homogeneous HMM with even 3 states, and gets much worse with 4 states. The multiple-block model gives approximately the same BIC as the homogeneous HMM, although it gives the BIC-optimal number of states as 4, in contrast to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal models are the best models tested by far; they both give the BIC-optimal number of states as 5, and they have similar goodness of fit.
However, this similarity is overstated due to the very large variation in BIC (over thousands of units) across the full range of models; there is a difference of approximately 80 BIC units, which would
normally be interpreted as an enormous difference in goodness of fit,
between the sinusoidal and quadratic models (both of which have 90 
parameters).

@@

HMMs are a widely used and flexible tool for modeling animal movement
behaviour; we need to work harder to make sure they are both appropriately
complex and biologically interpretable. 
With the increasing volumes of movement data available,
ecologists who naively use traditional homogeneous HMMs
and standard information-theoretic criteria
to estimate the number of behavioural states will generally
overfit their data, in the sense of ``discovering'' large
number of states that are difficult to interpret biologically.

On a broad spectrum, it really depends on what kind of question 
that is being answered. On one side of the spectrum, if the goal is 
to identify states, it might be sufficient to use a 
simple/traditional HMM model and pre-specify the number of states and, post hoc, 
match Viterbi-based states estimates with environmental variation 
\cite{fryxell_multiple_2008}. In our panther study, it is really hard to interpret multiple latent states biologically.
We have no way of knowing what panthers are actually
thinking (it is certainly more complex than being in one of a small
number of discrete latent states); 
we don't know the ``true'' number of latent states,
nor are we able to observe them directly, although
incorporating additional direct observations of 
behaviour (if available) can at least partially
address this problem \cite{fryxell_multiple_2008}.
Three distinct
movement states seem biologically interpretable for Florida panthers
according to van de Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
states, intermediate step length a foraging state, and long
step length a traveling state.

On the other side of the 
spectrum, if the goal of interest is to make predictions (out of sample), 
it might be better to fit a covariate-dependent model so that we can explicitly
model the switching process. In that case, fitting a covariate-dependent model is
better for out of sample prediction because Viterbi can only estimate state 
occupancy if observed movements are available (within sample predictions).

Finally, if we want to estimate the number of states, BIC is not necessarily 
good for estimation of number of states \cite{biernacki2000assessing}, 
but it can be useful as an approximate upper limit estimate. This is a common problem when using BICs, thus, the "knee point" (jump point) was proposed for BIC in partitioning based clustering via angle-based method \cite{zhao_knee_2008}. A similar method was used with likelihood \cite{dean2012behavioural} where both are applicable in our case, and suggest fewer number of states, but we do not know the reliability when fitting more complex models (IA. the shape of the BIC curve can be affected by the non-linear increase of parameters with every additional state).

Incorporating temporal heterogeneity in animal movement is one step in
the right direction, but much remains to be done. 
Our model neglects other predictors, such as sex, habitat type or
location with respect to environmental features such as roads,
that can potentially improve goodness of fit and predictions and 
further reduce the estimated number of states. While adding
more covariates is in principle straightforward using existing
frameworks, including all possible biological complexities in 
a HMM with state-dependent transitions may rapidly become intractable
in terms of both computational time and complexity of choosing
among possible reduced models and numbers of states.
Better diagnostic procedures and tests are needed:
these can both test overall goodness-of-fit \cite{potts_generalized_2014} 
and, more importantly, localize fitting problems to particular aspects of the
data so that models can be constructed without needing to
include all possible features of interest. Because there are a huge number of potential complexities that can be added to movement models (e.g. spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.

>>>>

Model complexity and the number of parameters increase as the number of latent states increase.
For a fixed number of states homogeneous FMMs are simplest, followed by homogeneous HMMs and finally by FMMs and HMMs incorporating temporal heterogeneity. In general, the number of free parameters in an HMM is the sum of the number of free parameters for each of the three model components (initial states, transition probabilities, and emissions). Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$. Then the number of free parameters of an HMM is:
[\emph{Initial states}] $k_{i}\cdot (n-1)$ 
+ [\emph{Transition probabilities}] $k_{t}\cdot n\cdot (n-1)$ 
+ [\emph{Emission parameters}] $k_{e}\cdot n$.
As the number of states increases, the number of free parameters in (homogeneous or heterogeneous) FMMs and time-homogeneous HMMs increases linearly, whereas for HMMs with temporal heterogeneity (or covariate-dependent transitions more generally) the number increases quadratically.

@@

We used the {\tt depmixS4} package \cite{visser2010depmixs4}
to fit covariate-dependent transition HMMs, 
simulate states and step lengths using the
estimated parameters, and estimate the most likely
states with the Viterbi
algorithm.

@@

We used simulations to predict expected hourly step lengths and
autocorrelation functions (ACF).  While the computation of expected
step length and ACF is straightforward for FMMs, and feasible for
homogeneous HMMs, the interaction between the geometric dwell time
within each state and the temporally varying interaction probabilities
makes it infeasible for more complex models.  We used this approach to
validate our models, comparing our simulated predictions with the
observed movements.  The more usual approach, generating predictions
from the expected step lengths conditional on the most likely state
sequence predicted by the Viterbi algorithm or pseudo-residuals \cite{zucchini_hidden_2009,langrock_flexible_2012}, is problematic
because the predictions by these methods rely on the observed data. 
This approach is useful to predict
missing data in the observation sequence, but because it is
conditional on the observed values, it can not reliably evaluate
goodness of fit for HMM models with
different degrees of structural complexity.

@@

%% fig 1 goes somewhere here
The simulation experiment supports our hypothesis that homogeneous transition HMMs can overestimate the number of hidden states (Figure~1). Heterogeneous transition models can always predict the correct number of states (in 100/100 simulations, BIC correctly identifies $n=2$ as the number of states) whereas the temporally homogeneous models overestimate the number of states (the correct value, $n=2$, is chosen most often, but in fewer than half of the simulations; values up to $n=5$ are frequently chosen).

\bmb{caption needs a little work}

@@

The BIC-optimal number of states for time-homogeneous models is
consistent with van de Kerk et al.'s \cite{kerk2015hidden}
results. For time-homogeneous models, the Weibull-wrapped-Cauchy
\cite{kerk2015hidden}, Weibull-von Mises, and log Normal without
turning angles all identify the same BIC-optimal number of
states. While the number of states identified
by homogeneous-HMM models varies according to the emissions
distributions chosen, ranging from $n=5$ for Weibull
steps alone to $n=7$ for the log Normal-von Mises emissions
model, the number of states identified
by heterogeneous-HMM models
is consistent among emissions models ($n=5$: Figure 2).

@@

Models with temporal heterogeneity provide better fits to the data
(lower BIC) than homogeneous models in both FMM and HMM frameworks,
but time-homogeneous HMMs are better than FMMs with sinusoidal
temporal heterogeneity (Figure 3). Turning to the temporally
heterogeneous HMMs (Figure 3, right panel), we see that the model with
different transition probabilities for each hour of the day (HMM +
THhourly) is overparameterized; it underperforms homogeneous HMM with
even 3 states, and gets much worse with 4 states. The multiple-block
model gives approximately the same BIC as the homogeneous HMM,
although it gives the BIC-optimal number of states as 4, in contrast
to 6 for the homogeneous HMM.  Finally, the quadratic and sinusoidal
models are the best models tested by far; they both give the
BIC-optimal number of states as 5, and they have similar goodness of
fit.  However, the similarity between the quadratic 
and sinusoidal models may be overstated in Figure~3 due to the very large
variation in BIC (over thousands of units) across the full range of
models; the best-fit sinusoidal ($n=5$) model is approximately 80 BIC units
better than the best quadratic model (also $n=5$), which
would normally be interpreted as an enormous improvement in goodness of
fit (both models have 90 parameters).

@@

HMMs are a widely used and flexible tool for modeling animal movement
behaviour; we need to work harder to make sure they are both appropriately
complex and biologically interpretable. 
With the increasing volumes of movement data available,
ecologists who naively use traditional homogeneous HMMs
and standard information-theoretic criteria
to estimate the number of behavioural states will generally
overfit their data, i.e. they will ``discover'' large
number of states that are difficult to interpret biologically.

As usual, the appropriate approach depends on the goal of the
analysis.  If ecologists simply want to identify states
and associate them with environmental characteristics, it might be
sufficient to use a simple (homogeneous) HMM model, pre-specifying the
number of states to a biologically sensible value, and then match
\emph{post hoc} Viterbi estimates of state occupancy with
environmental variation in space and time \cite{fryxell_multiple_2008}.

%% In our panther
%% study, it is really hard to interpret multiple latent states
%% biologically.  We have no way of knowing what panthers are actually
%% thinking (it is certainly more complex than being in one of a small
%% number of discrete latent states); we don't know the ``true'' number
%% of latent states, nor are we able to observe them directly, although
%% incorporating additional direct observations of behaviour (if
%% available) can at least partially address this problem
%% \cite{fryxell_multiple_2008}.  Three distinct movement states seem
%% biologically interpretable for Florida panthers according to van de
%% Kerk et al. \cite{kerk2015hidden}: Short step length suggests resting
%% states, intermediate step length a foraging state, and long step
%% length a traveling state.

On the other hand, if the goal of analysis is to make out-of-sample
predictions about animal behaviour, such as in a management context,
it is necessary to fit a covariate-dependent model that
explicitly incorporates the switching process.
While the Viterbi algorithm can be applied to
work backward from observed behaviour
to variations in state occupancy with environmental conditions
even when using a homogeneous model,
a homogeneous model 
can never \emph{predict} behaviour that varies 
with environmental conditions.

Finally, if our goal is actually to estimate the number of discrete
behavioural modes underlying observed behaviour --- keeping in mind
that these discrete behavioural states are certainly an oversimplified
representation of animals' real internal states --- then, as we have
shown above, covariate-dependent models will generally be required to
avoid overestimating the number of states.  We note that researchers
in cluster analysis (of which HMMs are a special case) have shown that
the technical conditions required for BIC to apply may be violated
\cite{biernacki2000assessing}; however, BIC can be useful as an
approximate \emph{upper} limit on the number of states. Various
solutions to this problem have been proposed, including the
``integrated classification likelihood''
\cite{biernacki2000assessing}, as well as a simpler ``knee point''
method \cite{zhao_knee_2008} that looks for the cluster size that
corresponds to the largest change in BIC rather than to the smallest
overall BIC.  \cite{dean2012behavioural} took a similar approach, but
based on the log-likelihood curve rather than the BIC.
However, at least in our simulations the BIC does correctly identify
the number of states when appropriate heterogeneity is included
in the model; the best practical criterion for identifying 
numbers of states underlying animal movement data is an open question.

\bmb{mike, can we take a quick try at computing ICL for our data/sims?}

\begin{equation*}
\begin{split}
ICL(m,K) & = \underbrace{\max_{a} \log f(x|\bar z, m, K, a)}_{\textrm{max log-likelihood}} - \frac{\lambda_m,K}{2} \log n \\
 & + \log \Gamma(K/2) + \sum_{k=1}^{K} \log \Gamma(\bar n_k + 1/2) \\
 & - K \log \Gamma(1/2) - \log \Gamma(n + K/2)
\end{split}
\end{equation*}

($K$=number of clusters; $n$=number of obs.; $\lambda_{m,K}$ is ``the
number of free components in $a$'' (i.e., number of free parameters:
p. 8); $\bar n_k$ is, I think, the number of data points which have
their max. probability of being in cluster $k$ [i.e., we should be
  able to get this at least approximately from Viterbi, although I'm
  not sure whether it's more appropriate to use Viterbi or an
  ``instantaneous'' max probability]). The {\tt rebmix} package has an
ICL criterion for its particular objects ({\tt
  sos::findFn("\{integrated classification likelihood\}"}), but I
don't know if it'll be useful.  I think Raftery and Fraley also have a
modified IC for clusters?

We have shown the value of incorporating temporal heterogeneity in 
animal movement models, but much remains to be done. 
Our model neglects other predictors --- such as sex, habitat type or
location with respect to environmental features such as roads --- %
that can potentially improve predictive accuracy and 
further reduce the estimated number of states. While adding
more covariates is in principle straightforward, 
including even a reasonable range of biological complexity in 
a HMM with state-dependent transitions rapidly becomes intractable
in both computational time and the complexity of model selection.
Better diagnostic procedures and tests are needed:
these can both test overall goodness-of-fit \cite{potts_generalized_2014} 
and, more importantly, localize fitting problems to particular aspects of the
data so that models can be constructed without needing to
include all possible features of interest. Because there are a huge number of potential complexities that can be added to movement models (e.g. spatial/temporal/among-individual heterogeneity; effects of conspecific attraction or avoidance; memory or cognitive effects), each with associated costs in researcher and computational effort, such diagnostic plots are invaluable.
